A Gentle Introduction into Structural Causal Models

### 2021-06-11 paper_scm.Rmd
recent v

**Abstract** The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling. Probabilistic specifications cast a model based on conditional probabilities. SCMs cast a model based on  assymetric assignments and extend probabilistic models by specifying the entire data generating process rather than solely utilizing conditional probabilities. 
Another difference between these models is their ability to address different queries such as *predictions*, *interventions* and *counterfactuals*. 
These queries are part of Pearl's causal hierarchy (2009).
Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*. 
I compare the feasibility of addressing these queries and undertaking respective actions for both specifications.
To contextualize SCMs within the field of causality, I also discuss the role of time in causality.
This paper uses various directed acyclic graphs to highlight the differences in these modelling approaches.
The insights of this paper can be used as a baseline for subsequent research on structural causal models.

@@

To accommodate existing literature, this paper provides a gentle introduction to SCMs, focusing on the underlying assumptions we make when building a structural causal model. 
The aim of this paper is to provide an example based, inutitive understanding of different concepts within causality using manifold elements from core literature on causality.

The rest of the paper is structured as follows:
Section 2 introduces the assumptions in causal modelling.
Section 3 addresses Pearls Causal Hierachy.
Section 4 focuses on the intersection of SCMs and the perception of time. 
Section 5 concludes.

@@

@@

We can craete graphical models based on various different algorithms.

@@

Graphical models developed a mathematical langauge to assess certain conditions. 
One pivotal feature is the d-separation [@judea2010introduction]:

**Definition 2** ( $d$ -separation) A set $S$ of nodes is said to block a path $p$ if either (i) $p$ contains at least one arrow-emitting node that is in $S$, or $($ ii $)$ p contains at least one collision node that is outside $S$ and has no descendant in $S .$ 
If $S$ blocks all paths from $X$ to $Y$, it is said to " $d$ -separate $X$ and $Y, "$ and then, $X$ and $Y$ are independent given $S$, written $X \perp \!\!\! \perp Y \mid S .$

Select subset of variables with backdoor criterion:

## Independence

(a) possible to undertake local intervention -> change f(x), regardless of f(y|x)
(b) these components are autonomous objects -> set of autonomous equations

Independence of Noise,

Noise independent so problematic behaviour and genetic code should be independent.

## Time

 why time is ignored.

@@

Pearl [@bareinboim2020pearl; @pearl2009causality] introduced the hierachy of causation to categorize different statistical and causal tools. 
The hierarchy of causation contains (see table 1) three levels. 
The higher the method on the hierarchy, the more information the method requires.
This section discusses each of these methods and their respective advantages.

\footnotesize
 |Method          | Action |  Example | Usage | 
|------------------|-------------|--------------------|-------------------|
| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO

Table: Pearls Hierachy of Causation (2009)

\normalsize

@@

```{tikz, notwell,  fig.cap ="Probabilistic Model",  fig.align="center", echo =F, fig.show="hold", out.width="50%" }
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\node[my node] (S){S};
\node[my node,right =1 cm of S](G){G};
\node[my node] at ($(S)!0.5!(G)-(0pt,1.5cm)$) (C) {C};
\draw[->] (S) -- (C);
\draw[->] (G) -- (C);
\end{tikzpicture}
```

@@

One extension of policy intervention is off-policy intervention. 
**Off-policy intervention** models different intervention that is not in our historical data.
To accommodate the lack of data, we can estimate the intervention either based on model-based (e.g. regression),model-free estimates (propensity scores), or a mixture of these methods. 
For further information, see e.g. @oberst2019counterfactual or [@creager].

$do(t)$ -> replace function $T:=f_T(\pi)$ with different conditional probability (or constant) [@pearl2009causality].

**Counterfactuals:**

@@

$$
\begin{array}{|l|l|l|l|l|}
\hline \text { model } & \begin{array}{l}
\text { IID setting }
\end{array} & \begin{array}{l}
\text { changing } \\
\text { distributions } 
\end{array} & \begin{array}{c}
\text { counter- } \\
\text { factual } \\
\text { questions }
\end{array} & \begin{array}{l}
\text { physical } \\
\text { insight }
\end{array} \\
\hline \begin{array}{l}
\text { mechanistic } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
\hline \begin{array}{l}
\text { structural } \\
\text { causal model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
\hline \begin{array}{l}
\text { causal } \\
\text { graphical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
\hline \begin{array}{l}
\text { statistical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
\hline
\end{array}
$$

Table: Source: @peters_elements_2017

@@

Structural causal models are flexible simulators to disentangle causality for manifold different queries.
There are many advantages of structural causal models:
(1) We are able to model latent fators forcing us to reconsider existing assumptions about the relationship in our data.
(2) Further, we get a underlying graphical representation including a mathematical language for this graphical systems to test causal assumptions that are otherwise untestable.
(3) Additionally, one is able to model queries beyond mere association going as far as dealing with hyptoethical situations. 
Simultaneously,

>>>>

**Abstract** The interest in understanding relationships of variables beyond co-occurrence has increased the popularity of causal modelling. 
Probabilistic specifications cast a model based on conditional probabilities. 
SCMs cast a model based on  asymmetric assignments and extend probabilistic models by specifying the entire data generating process rather than solely utilizing conditional probabilities.
A SCM is build on a number of different assumptions namely the independence of noise, autonomy of mechanisms, causal sufficiency and the causal markov condition.
I also discuss the role of time in causality and how differential equations enable us to model SCMs similar to a mechanical mdoel.
Another difference between thes SCM and other model specifications is the ability  of SCMs to address different queries such as *predictions*, *interventions* and *counterfactuals*. 
These queries are part of Pearl's causal hierarchy (2009).
Pearl matches these queries with their respective actions namely *observing*, *doing* and *imagining*. 
I compare the feasibility of addressing these queries.
The insights of this paper can be used as a baseline for subsequent research on structural causal models.

@@

To accommodate existing literature, this paper provides a gentle introduction to SCMs, focusing on the underlying assumptions we make when building a structural causal model.
Specifically, this paper focuses on the independence of noise, independence of mechanisms, causal sufficiency, the role of time and the causal markov condition.
Additionally, this paper provides an overview of various other statistical and causal methods within the grand scheme of pearls causal hierachy.
I accomodate this outlook with various considerations for the usage of SCMs. 
The aim of this paper is to provide an example based, intuitive understanding of different concepts within causality using manifold elements from core literature on causality.

The rest of the paper is structured as follows:

@@

## Mathematical Components

@@

## Independence of Noise and Mechanism

- possible to undertake local intervention -> change f(smoking), regardless of f(cancer|smoking)
- these components are autonomous objects -> set of autonomous assignments

**Independence of Noise:**

The way we view noise in SCMs differs from the classical view in regression analysis.
Therefore, I will briefly clarify essential differences. 
Noise variables are our latent variables which are the parent nodes of our child nodes of interest [@hardtrecht].
In classical regression analysis, such as an ordinary least squares regression, we employ the Gauss Markov Assumptions. 
The exogeneity assumption, one of the five Gauss Markov Assumptions, suggests that the error terms are uncorrelated with our features [@wooldridge2010econometric].
For the SCM, the error terms are driving our features and a pivotal component of the model specification [@hardtrecht].
To ensure that we are correctly specifying mechanisms and our underlying model structure, one needs to ensure independence of these noise terms. 
As mentioned in the definition for SCMs, we would mathematical notate this independence as follows:

$N_{F} \perp \!\!\! \perp N_{C}$

Scholarship in causality established the causal sufficiency condition:

By ensuring that this condition holds, we can easier disentangle underlying common causes, avoiding a violation of the independence of our noise terms.
Independence of noise is one form of causal sufficiency [@peters_elements_2017].

There are two related dimensions to consider namely the informational aspect of independence  and implications for  modularity. 
If the independence conditions holds, and the cause and effect variables are independent, the cause and effect variable do not contain information about each other. 
Modularity [@pearl2009causality] describes the advantage of being able to treat these variables as distinct modules.
This means, even if we have a change in one variable (cause or effect), we can disentangle our variable as unique modular components.
Especially looking at machine learning, domain shift and covariate shift are problematic for classical tools. [@peters_elements_2017]
By using these independence assumptions in SCMs, we can accommodate these settings.

Looking at our smoking example:

# Directed Acyclic Graphs

@@

Graphical models developed a mathematical langauge to assess certain conditions.

We can create graphical models based on various algorithms.

We can obtain a graphical model based on (a single) observational dataset through conditional independence testing.
The graphical approach to recover a graph from an observaational dataset requires two assumptions namely jointly independent noise terms and faithfulness. 
Intuitively speaking, this method examines how noise spreads [@peters_elements_2017]. 
Two other methods are ICM^[The intuition here is that the noise terms cultivate the footprint of the assignments. For more information see: @shajarisales2015telling] and using an additive noise model^[This method uses regression and conditional independence testing to disentangle graphical structures. For further information see @mooij2016distinguishing].

[@peters_elements_2017] provide a taxonomy of different methods (see table 1)

$$\text{Table 1 Source: Peters et al. (2017) , Modelling Taxonomy}$$
$$
\begin{array}{|l|l|l|l|l|}
\hline \text { model } & \begin{array}{l}
\text { IID setting }
\end{array} & \begin{array}{l}
\text { changing } \\
\text { distributions } 
\end{array} & \begin{array}{c}
\text { counter- } \\
\text { factual } \\
\text { questions }
\end{array} & \begin{array}{l}
\text { physical } \\
\text { insight }
\end{array} \\
\hline \begin{array}{l}
\text { mechanistic } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} \\
\hline \begin{array}{l}
\text { structural } \\
\text { causal model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} \\
\hline \begin{array}{l}
\text { causal } \\
\text { graphical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{Y} & \mathrm{N} & \mathrm{N} \\
\hline \begin{array}{l}
\text { statistical } \\
\text { model }
\end{array} & \mathrm{Y} & \mathrm{N} & \mathrm{N} & \mathrm{N} \\
\hline
\end{array}
$$

In the table we can see that traditional statistical models fair well in IID settings, where our observations are independent and identically distributed.
But once we encounter a changing distribution such as an covariate shift or a domain shift, these statistical models are not suitable anymore.
Statistical models are also unable to answer any counterfactual questions about hypothetical settings outside of our observational data and are unable to provide physical insights because we do not model time as an explicity. 
Causal graphical models are able to deal with changing distributions and fair well in the IID setting. 
Nevertheless, they are also unable to answer counterfactual questions on its own and cannot provide physical insights. 
SCMs are able to deal with the first three queries, but largely ignore time.
By commonly making assumptions such as that the effect is acyclic, SCMs simplify time and treat focus on equilbrium states.
The mechanistic view, where we model time as an explicit factor, enables us to answer all those queries. 
Note that the growing complexity makes it very challenging to implement but relevant for natural sciences where time is pivotal to disentangle relationships [@mooij2013ordinary; @peters_elements_2017].

@@

Pearl [@bareinboim2020pearl; @pearl2009causality] introduced the hierachy of causation to categorize different statistical and causal tools. 
The hierarchy of causation contains (see table 1) three levels, where the high methods on the hierarchy, the more information the method requires.
This section discusses each of these methods and their respective advantages.

$$\text{Table 2 Source: Pearl (2009) , Hierarchy of Causation}$$ 
\footnotesize
|Method          | Action |  Example | Usage | 
|------------------|-------------|--------------------|-------------------|
| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO            
\normalsize

@@

```{tikz,fig.cap="Probabilistic Model" ,fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (p){S};
\node[my node, fill=gray!30] at ($(p)!0.5!(p)-(0pt,1.5cm)$) (G) {G};
\node[my node,right = 1 cm of G, fill=gray!30](C){C};
\draw[->] (p) -- (C);
\draw[->] (G) -- (C);
\end{tikzpicture}
```

@@

One way of the most prominent ways to estimate casual effects of our treatment is the average treatment effect.
The average treatment effect looks at the difference between treatment and control group.
For the case of simplicity in our example, we want to look at whether smoking at all has an impact on lung cancer.
Mathematically, this could as follows:

\begin{equation}
f(C|S=1) - f(C|S=0)
\end{equation}

where $S = 1$ means the participant smokes and $S=0$ means the participant does not smoke.
Furthermore, it is crucial that we hold all other factors constant so looking at our example, we would compare people with the same genetic features to truely isolate the effect of smoking.
The ATE is a population level effect [@hardtrecht].

**Confounders**

Confounding is the circumstance where the observational information (conditional probabilities in our data) and intervention are different.
Essentially, we want to express our do-intervention based on conditional probabilities, but have no direct information in our data.
To still estimate a conditional probability, we can use a mixture of existing conditional probabilities and estimate the desired probability based on the adjustment formula.
For further information, see  [@hardtrecht].

## Counterfactuals

@@

## Considerations

**If you...**

* ..only have conditional probabilities 
* ..are only interested in association 
* ..have very limited knowledge of your actual underlying structure

$\rightarrow$ you should probably go for a probabilistic model, association based model or a model based on conditional probabilities

**But if you**

* ..have an understanding of underlying relationships in the data
* ..are dealing with high stake settings 
* ..want to simulate different interventions

$\rightarrow$ you could probably go for a SCM

@@

Structural causal models are flexible simulators to disentangle causality for manifold different queries.
There are many advantages of structural causal models:
(1) We are able to model latent factors forcing us to reconsider existing assumptions about the relationship in our data.
(2) Further, we get a underlying graphical representation including a mathematical language for this graphical systems to test causal assumptions that are otherwise untestable.
(3) Additionally, one is able to model queries beyond mere association going as far as dealing with hyptoethical situations.

Apart from the assumptions, causal graphics can help reseachers avoiding mistakes.
To determine this causal graph, we can use conditional independence testing and assess assumptions about faithfulness and jointly independent noise terms.

Looking at the causal hierachy, structural causal modelling enables us to compute interventions efficienctly given certain condition hold. 
These conditions defintely not always hold true in every instance. 
Especially looking at the
